---
title: "Cerebellar layer quantification — example analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This notebook demonstrates the full analysis pipeline using **simulated data**.
Replace the `simulate_cerebellar_image()` calls and the genotype mapping with your
own `.tif` files and sample metadata to run on real data.

The pipeline produces depth-resolved intensity profiles across the cerebellar
cortex and compares them between two experimental groups.

## Install / load packages

```{r packages, eval=FALSE}
install.packages(c(
  "dplyr", "tidyr", "ggplot2", "patchwork",
  "stringr", "zoo", "pracma",
  "RColorBrewer", "ggrepel", "ggsignif",
  "lme4", "lmerTest"
))

if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("EBImage")
```

```{r load_packages}
library(EBImage)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(stringr)
library(zoo)
```

## Source helper functions

All helper functions live in the same directory as this notebook.

```{r source_functions}
source("align_folia_images_connect.R")
source("classify_connectivity_segments.R")
source("elg_igl_normalize.R")
source("pia_prunner.R")
source("plot_layer_boundary2.R")
```

---

## Part 1 — Simulate example images

The function below creates synthetic three-channel TIFF files that mimic the
structure of a P7 cerebellar cortex cross-section:

- **DAPI** (channel 1): elevated in EGL (rows 50–120) and IGL (rows 280–380)
- **p27** (channel 2): peaks at the EGL/ML boundary (~row 80), low elsewhere
- **NeuN** (channel 3): low in EGL, peaks in IGL

Row 0 corresponds to the pial surface; higher row numbers = deeper layers.

In real data, replace this section with your own `.tif` files in a directory.

```{r simulate_data}
pixel_width <- 0.5119  # micrometers per pixel (adjust for your objective)

simulate_cerebellar_image <- function(n_rows    = 500,
                                      n_cols    = 600,
                                      noise_sd  = 0.03,
                                      pia_offset = 0) {
  rows <- seq_len(n_rows) + pia_offset

  # Gaussian bump helper
  bump <- function(x, centre, width, height)
    height * exp(-0.5 * ((x - centre) / width)^2)

  # --- DAPI: two dense nuclear layers (EGL + IGL) ---
  dapi <- bump(rows,  85, 30, 0.8) +   # EGL
          bump(rows, 330, 50, 0.9) +   # IGL
          0.15 + rnorm(n_rows, 0, noise_sd)

  # --- p27: marks post-mitotic cells, peaks at EGL/ML boundary ---
  p27  <- bump(rows,  80, 20, 0.9) +
          bump(rows, 160, 40, 0.3) +   # faint ML signal
          0.05 + rnorm(n_rows, 0, noise_sd)

  # --- NeuN: mature neurons, peaks in IGL / Purkinje layer ---
  neun <- bump(rows, 200, 25, 0.5) +   # Purkinje layer
          bump(rows, 320, 55, 0.9) +   # IGL
          0.05 + rnorm(n_rows, 0, noise_sd)

  # Clamp to [0, 1] and build an EBImage multi-frame object
  clamp <- function(x) pmax(0, pmin(1, x))

  # Each frame is n_rows × n_cols; columns are uniform (same profile per column)
  make_frame <- function(profile) {
    mat <- matrix(clamp(profile), nrow = n_rows, ncol = n_cols)
    mat + matrix(rnorm(n_rows * n_cols, 0, noise_sd / 3),
                 nrow = n_rows, ncol = n_cols)
  }

  img <- EBImage::combine(
    Image(make_frame(dapi)),
    Image(make_frame(p27)),
    Image(make_frame(neun))
  )
  img
}

# Write simulated images to a temporary directory -------------------------
sim_dir <- file.path(tempdir(), "sim_images")
dir.create(sim_dir, showWarnings = FALSE)

set.seed(42)

mouse_info <- data.frame(
  mouse_id  = c("M1", "M2", "M3", "M4", "M5", "M6"),
  genotype  = c("WT", "WT", "WT", "KO", "KO", "KO"),
  pia_noise = c(0, 3, -2, 1, -3, 2)   # small random pia offsets per animal
)

for (i in seq_len(nrow(mouse_info))) {
  for (fol in c("folium_a", "folium_b")) {
    img  <- simulate_cerebellar_image(pia_offset = mouse_info$pia_noise[i])
    fname <- file.path(sim_dir,
                       sprintf("%s_%s_%s.tif",
                               mouse_info$genotype[i],
                               mouse_info$mouse_id[i],
                               fol))
    writeImage(img, fname)
  }
}

cat("Simulated images written to:", sim_dir, "\n")
list.files(sim_dir)
```

---

## Part 2 — Load images and compute row statistics

`align_folia_images_connect()` loads every `.tif` in the directory, computes
row-wise statistics per column segment, then aligns all image strips to the one
with the highest average cross-correlation to the rest.

**Real data:** point `sim_dir` at your own directory of pre-cropped TIFF strips.

```{r load_images}
aligned_data <- align_folia_images_connect(
  sim_dir,
  frame_names          = c("DAPI", "p27", "NeuN"),
  reference_frame      = "DAPI",
  pixel_segment_length = 200
)

cat("Rows in aligned data:", nrow(aligned_data), "\n")
cat("Channels:", unique(aligned_data$Frame), "\n")
cat("Files loaded:", length(unique(aligned_data$filename)), "\n")
```

---

## Part 3 — Attach genotype metadata

Parse the animal identifier out of the filename and join the genotype table.

**Real data:** replace the `str_extract` pattern with whatever identifies your
animals in your filenames.

```{r genotyping}
# Extract mouse ID from filename  (pattern: GENOTYPE_MOUSEID_FOLIUM_segment_N)
aligned_data <- aligned_data |>
  mutate(
    mouse_id = str_extract(filename, "(?<=_)(M\\d)(?=_)"),
    genotype = str_extract(filename, "^(WT|KO)")
  )

cat("Genotype breakdown:\n")
aligned_data |>
  distinct(filename, genotype, mouse_id) |>
  count(genotype, mouse_id) |>
  print()
```

---

## Part 4 — Quality filtering

Keep only image strips that align well to the reference (MaxCorrelation > 0.5).

```{r qc_filter}
# Plot correlation distribution to choose threshold
aligned_data |>
  distinct(filename, Frame, MaxCorrelation) |>
  filter(Frame == "DAPI") |>
  ggplot(aes(x = MaxCorrelation)) +
  geom_histogram(binwidth = 0.05, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = 0.5, linetype = "dashed", colour = "red") +
  labs(title = "Cross-correlation quality filter",
       x = "Max correlation to reference image",
       y = "Number of image strips") +
  theme_minimal()

filtered_data <- aligned_data |> filter(MaxCorrelation > 0.5)
cat("Strips remaining after QC:", length(unique(filtered_data$filename)), "\n")
```

---

## Part 5 — Prune sparsely covered rows

Remove row positions that are not covered by at least 80% of image strips. This
trims the ends where only a few images contribute, which would otherwise produce
noisy averages.

```{r prune_rows}
min_relative_coverage <- 0.8

row_coverage <- filtered_data |>
  group_by(Row) |>
  summarise(n_files = n_distinct(filename),
            rel_coverage = n_files / max(n_files)) |>
  filter(rel_coverage >= min_relative_coverage)

pruned_data <- filtered_data |>
  inner_join(row_coverage, by = "Row")

cat("Row range after pruning:", range(pruned_data$Row), "\n")
```

---

## Part 6 — Shift to common origin

Find the local minimum of the average p27 profile (corresponds to the pial
surface / EGL peak transition) and shift all data so this point is at Row = 0.

```{r shift_origin}
# Average p27 profile across all strips
avg_p27 <- pruned_data |>
  filter(Frame == "p27") |>
  group_by(Row) |>
  summarise(AvgRowMean = mean(RowMean, na.rm = TRUE))

find_local_minimum <- function(data) {
  diffs <- diff(data$AvgRowMean)
  change_points <- which(diffs[-1] > 0 & diffs[-length(diffs)] < 0)
  centre <- nrow(data) / 2
  change_points[which.min(abs(change_points - centre))]
}

origin_row <- find_local_minimum(avg_p27)
cat("Origin row (p27 local minimum):", origin_row, "\n")

shifted_data <- pruned_data |>
  mutate(Row_shift = Row - origin_row,
         Row_micron = Row_shift * pixel_width)
```

---

## Part 7 — Split at origin and flip negative side

Each image strip contributes signal on both sides of the pial reference point.
We split the data at Row_shift = 0, flip the negative side to positive, and
stack both halves — doubling the number of profile replicates.

```{r split_halves}
neg_half <- shifted_data |>
  filter(Row_shift <= 0) |>
  mutate(Row_shift = abs(Row_shift),
         xpslit    = "negative")

pos_half <- shifted_data |>
  filter(Row_shift >= 0) |>
  mutate(xpslit = "positive")

split_data <- bind_rows(neg_half, pos_half)
```

---

## Part 8 — Per-strip intensity rescaling

Rescale each strip to its own 5th–95th percentile range. This corrects for
differences in overall staining intensity between sections and animals.

```{r rescale}
rescaled_data <- split_data |>
  group_by(Frame, filename, xpslit) |>
  mutate(
    RescaledIntensity = (RowMean - quantile(RowMean, 0.05, na.rm = TRUE)) /
      (quantile(RowMean, 0.95, na.rm = TRUE) -
         quantile(RowMean, 0.05, na.rm = TRUE)) * 100
  ) |>
  ungroup()
```

---

## Part 9 — EGL / IGL normalization

Normalize using the EGL (high) and molecular layer (low) as anatomical anchors.
This produces `RescaledIntensityInRange`, which is directly comparable across
experiments imaged at different times or on different microscopes.

```{r normalize}
# Need Row_shift_scale column name for egl_igl_normalize()
rescaled_data <- rescaled_data |> rename(Row_shift_scale = Row_shift)

normalized_data <- egl_igl_normalize(
  rescaled_data,
  DAPI_max_site = "igl",   # IGL is the dense nuclear layer at P7
  smooth        = TRUE
)
```

---

## Part 10 — Remove strips lacking IGL signal

Strips where the smoothed p27 peak in the first 150 pixels exceeds ~200% of
the normalized scale typically lack IGL tissue and would have distorted
normalization. Inspect the histogram and adjust the threshold for your data.

```{r remove_high}
max_egl <- normalized_data |>
  filter(Frame == "p27", Row_shift_scale <= 150) |>
  group_by(filename, xpslit) |>
  mutate(smoothed = zoo::rollmean(RescaledIntensityInRange, k = 10, fill = NA)) |>
  summarise(MaxSmoothed = max(smoothed, na.rm = TRUE), .groups = "drop")

ggplot(max_egl, aes(x = MaxSmoothed)) +
  geom_histogram(bins = 30, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = 200, linetype = "dashed", colour = "red") +
  labs(title = "Peak p27 intensity — filter strips above threshold",
       x = "Max smoothed p27 (first 150 px)",
       y = "Count") +
  theme_minimal()

high_tracks <- max_egl |> filter(MaxSmoothed > 200) |> select(filename, xpslit)
normalized_data <- anti_join(normalized_data, high_tracks,
                              by = c("filename", "xpslit"))
cat("Strips after IGL filter:", length(unique(normalized_data$filename)), "\n")
```

---

## Part 11 — Detect and trim pial boundary

`pia_prunner()` finds the position of the pial surface in each strip from the
joint drop in p27 and NeuN signal. The output `peak_position` is the pixel
offset to subtract so that position 0 = true pial surface.

```{r pia_trim}
pia_peaks <- pia_prunner(normalized_data)

pia_aligned <- normalized_data |>
  left_join(pia_peaks |> select(filename, xpslit, peak_position),
            by = c("filename", "xpslit")) |>
  mutate(
    AlignedRow       = Row_shift_scale - peak_position,
    AlignedRow_micron = AlignedRow * pixel_width
  )

# Quick sanity check: all strips aligned?
pia_aligned |>
  filter(Frame == "p27") |>
  ggplot(aes(x = AlignedRow_micron, y = RescaledIntensityInRange,
             group = paste0(filename, xpslit))) +
  geom_line(alpha = 0.15, colour = "black") +
  stat_summary(aes(group = genotype, colour = genotype),
               fun = mean, geom = "line", linewidth = 1) +
  xlim(0, 200) +
  labs(title = "p27 profiles after pial alignment",
       x = "Distance from pial surface (µm)",
       y = "Normalized p27 intensity") +
  theme_minimal()
```

---

## Part 12 — Build metadata table for biological replicates

The statistical tests in `plot_layer_boundary2()` average over sections from
the same animal before comparing groups. We need a table that maps each
filename back to its mouse ID.

**Real data:** construct `extra_anno` from your own filename structure.

```{r metadata}
extra_anno <- pia_aligned |>
  distinct(filename, xpslit, mouse_id, genotype) |>
  rename(Mouse = mouse_id)
```

---

## Part 13 — Final publication figure

`plot_layer_boundary2()` assembles three panels:

1. Individual strip profiles (faint) with group averages coloured by genotype
2. Average intensity difference (KO − WT) across position
3. Violin plots at the positions of maximum and minimum group difference,
   with biological replicate means overlaid and a t-test p-value

```{r final_figure, fig.height=9, fig.width=5}
fig <- plot_layer_boundary2(
  data             = pia_aligned,
  sel_frame        = "p27",
  x_col_name       = "AlignedRow_micron",
  y_col_name       = "RescaledIntensityInRange",
  x_min            = 0,
  x_max            = 150,
  y_min            = 0,
  y_max            = 150,
  comparison_order = c("WT", "KO"),
  biological_rep   = "Mouse",
  metadata         = extra_anno,
  stat             = "ttest"
)

print(fig)
```

---

## Pixel width reference

| Objective | Pixel width (µm) |
|-----------|-----------------|
| 20×       | 0.5119          |
| 10×       | 1.0238 (scale ×2 relative to 20×) |

When combining datasets acquired on different objectives, multiply the 10×
`Row_shift` values by 2 before merging so that all positions are in equivalent
20× pixel units, then convert to microns.

---

## Session info

```{r session_info}
sessionInfo()
```
