---
title: "Cerebellar layer quantification — example analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This notebook walks through the full pipeline using two real example images
included in `example_data/`. One image is from a WT animal (20x, dataset 1)
and one from an Ezh2 cKO animal (10x, dataset 2).

A complete between-genotype comparison requires multiple animals and sections
per group — the structure here is identical; just point the directory paths at
your full image sets and fill in the genotype mapping for your filenames.

Each `.tif` is a three-channel image: the pial surface is at the top (low row
index) and deeper layers are toward the bottom. Channel order differs between
the two datasets (see the `frame_names` argument below).

## Install / load packages

```{r packages, eval=FALSE}
install.packages(c(
  "dplyr", "tidyr", "ggplot2", "patchwork",
  "stringr", "zoo", "pracma",
  "RColorBrewer", "ggrepel", "ggsignif",
  "lme4", "lmerTest"
))
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install("EBImage")
```

```{r load_packages}
library(EBImage)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(stringr)
library(zoo)
```

```{r source_functions}
source("align_folia_images_connect.R")
source("classify_connectivity_segments.R")
source("elg_igl_normalize.R")
source("pia_prunner.R")
source("plot_layer_boundary2.R")
```

---

## Quick look at the example images

```{r view_images, fig.height=4, fig.width=9}
img_wt  <- readImage("example_data/2018_05_22_s3_5_p27-0006_fused_crop_3_4.tif")
img_ko  <- readImage("example_data/s6_E_p27_cor_merge_crop-5_6B.tif")

cat("WT image  — dimensions (rows × cols × channels):", dim(img_wt), "\n")
cat("KO image  — dimensions (rows × cols × channels):", dim(img_ko), "\n")

# Display channel 1 (p27) for each image side by side
par(mfrow = c(1, 2), mar = c(2, 2, 2, 1))
display(img_wt[,,1], method = "raster", title = "WT — p27 (ch 1)")
title("WT — p27")
display(img_ko[,,1], method = "raster")
title("KO — p27")
```

---

## Part 1 — Load images and compute row statistics

`align_folia_images_connect()` loads every `.tif` in a directory and returns a
long-format data frame with row-wise intensity statistics per channel per
column segment. Multiple segments per image give replicate measurements across
the tissue width.

### Dataset 1 — 20x (pixel width: 0.512 µm)

Channel order in these files: `p27 | NeuN | DAPI`

```{r load_dataset1}
pixel_width_20x <- 0.5119  # µm per pixel

aligned_ds1 <- align_folia_images_connect(
  "example_data/",              # replace with your full directory
  frame_names          = c("p27", "NeuN", "DAPI"),
  reference_frame      = "DAPI",
  pixel_segment_length = 200
)

# Attach genotype — for the example image mouse 3 = WT
# In a full analysis, build this table from all filenames
aligned_ds1 <- aligned_ds1 |>
  mutate(
    mouse_id = "M3",
    genotype = "WT",
    dataset  = 1L
  )

cat("Rows:", nrow(aligned_ds1), "| Files:", n_distinct(aligned_ds1$filename), "\n")
```

> **For your own data:** replace `"example_data/"` with your image directory and
> build the genotype mapping by extracting mouse identifiers from the filenames.

---

## Part 2 — Quality filtering and row coverage

Keep only segments that align well to the reference image
(`MaxCorrelation > 0.5`), then remove row positions that are not covered by at
least 80% of segments.

```{r qc_filter}
filtered <- aligned_ds1 |> filter(MaxCorrelation > 0.5)

row_coverage <- filtered |>
  group_by(Row) |>
  summarise(rel_coverage = n_distinct(filename) / max(n_distinct(filename)),
            .groups = "drop") |>
  filter(rel_coverage >= 0.8)

pruned <- inner_join(filtered, row_coverage, by = "Row")
cat("Row range after pruning:", range(pruned$Row), "\n")
```

---

## Part 3 — Shift to common origin

Find the local minimum of the average p27 profile and shift all rows so that
this point (approximately the EGL/ML boundary) is at position 0.

```{r shift_origin}
avg_p27 <- pruned |>
  filter(Frame == "p27") |>
  group_by(Row) |>
  summarise(AvgRowMean = mean(RowMean, na.rm = TRUE), .groups = "drop")

find_local_minimum <- function(df) {
  d <- diff(df$AvgRowMean)
  cps <- which(d[-1] > 0 & d[-length(d)] < 0)
  cps[which.min(abs(cps - nrow(df) / 2))]
}

origin_row <- find_local_minimum(avg_p27)
cat("Origin row:", origin_row, "\n")

shifted <- pruned |>
  mutate(Row_shift = Row - origin_row,
         Row_micron = Row_shift * pixel_width_20x)
```

---

## Part 4 — Split at origin and flip negative half

Split the data at Row_shift = 0, flip the negative side to positive, and stack
both halves. This doubles the number of profile replicates — one for each side
of the tissue strip relative to the reference point.

```{r split_halves}
neg_half <- shifted |> filter(Row_shift <= 0) |>
  mutate(Row_shift = abs(Row_shift), xpslit = "negative")
pos_half <- shifted |> filter(Row_shift >= 0) |>
  mutate(xpslit = "positive")

split_data <- bind_rows(neg_half, pos_half) |>
  rename(Row_shift_scale = Row_shift)
```

---

## Part 5 — EGL / IGL normalization

Normalize intensity using anatomical landmarks as anchors rather than
per-section percentiles. This makes values directly comparable across imaging
sessions and animals.

```{r normalize}
normalized <- egl_igl_normalize(
  split_data,
  DAPI_max_site = "igl",
  smooth        = TRUE
)
```

---

## Part 6 — Detect and trim pial boundary

`pia_prunner()` finds where p27 and NeuN both drop abruptly (the pial surface)
and returns a per-strip offset. After joining, `AlignedRow = 0` is the true
pial surface.

```{r pia_trim}
pia_peaks <- pia_prunner(normalized)

pia_aligned <- normalized |>
  left_join(pia_peaks |> select(filename, xpslit, peak_position),
            by = c("filename", "xpslit")) |>
  mutate(
    AlignedRow       = Row_shift_scale - peak_position,
    AlignedRow_micron = AlignedRow * pixel_width_20x
  )
```

---

## Part 7 — Plot intensity profiles

```{r plot_profiles, fig.height=5, fig.width=7}
pia_aligned |>
  filter(AlignedRow_micron >= 0, AlignedRow_micron <= 200) |>
  ggplot(aes(x = AlignedRow_micron, y = RescaledIntensityInRange,
             group = paste0(filename, xpslit))) +
  geom_line(alpha = 0.3, colour = "black") +
  facet_wrap(~Frame) +
  labs(
    title = "Intensity profiles after pial alignment",
    x     = "Distance from pial surface (µm)",
    y     = "Normalized intensity"
  ) +
  theme_minimal()
```

---

## Part 8 — Final combined figure (single genotype example)

`plot_layer_boundary2()` is designed for a two-group comparison. With a single
genotype the difference plot will be flat, but the call below shows the
correct structure. In a full analysis, load both genotypes into a single data
frame (following the same steps above) before calling this function.

```{r final_figure, fig.height=9, fig.width=5}
# Minimal metadata table mapping filename → mouse
extra_anno <- pia_aligned |>
  distinct(filename, xpslit) |>
  mutate(Mouse = "M3", genotype = "WT")

# To compare two genotypes, bind rows from both groups here and set
# comparison_order = c("WT", "KO") before calling plot_layer_boundary2().

plot_layer_boundary2(
  data             = pia_aligned,
  sel_frame        = "p27",
  x_col_name       = "AlignedRow_micron",
  y_col_name       = "RescaledIntensityInRange",
  x_min            = 0,
  x_max            = 150,
  y_min            = 0,
  y_max            = 150,
  comparison_order = c("WT", "WT"),   # replace with c("WT","KO") for real comparison
  biological_rep   = "Mouse",
  metadata         = extra_anno,
  stat             = "ttest"
)
```

---

## Notes on combining datasets from different objectives

When merging images acquired at different magnifications, scale the pixel
coordinates before combining:

| Objective | Pixel width (µm) | Scale factor relative to 20x |
|-----------|-----------------|------------------------------|
| 20×       | 0.5119          | 1×                           |
| 10×       | 1.0238          | ×2 (multiply `Row_shift` by 2) |

After scaling, all positions are in equivalent 20× pixel units. Convert to
microns using `pixel_width_20x = 0.5119`.

---

## Session info

```{r session_info}
sessionInfo()
```
